Reanalysis Step 4: Outcome Analysis with Three-Way Interaction

### 2025-08-19 reanalysis-3-propensity_scores.qmd
analysis audit

```{r diagnostics-loop}
# Generate diagnostic plots for selected imputations
for (i in imps_to_show) {
  cat("\n### Imputation", i, "\n\n")

  twang_model <- twang_results[[i]]$model

  cat("#### Convergence\n")
  plot(twang_model, plots = 1)

  cat("\n#### Weight Distribution\n")
  plot(twang_model, plots = 2)

  cat("\n#### Covariate Balance\n")
  plot(twang_model, plots = 3)

  cat("\n#### Weight Proportions\n")
  plot(twang_model, plots = 6)

  cat("\n#### Balance Summary Table\n")
  print(summary(twang_model))
}
```

>>>>

```{r diagnostics-loop}
#| results: "asis"

# Generate diagnostic plots for selected imputations
for (i in imps_to_show) {
  cat("\n### Imputation", i, "\n\n")

  twang_model <- twang_results[[i]]$model
  weights <- twang_results[[i]]$weights

  # Create dataset with weights for cobalt
  data_with_weights <- imputed_datasets[[i]]
  data_with_weights$weights <- weights

  cat("#### Convergence (Iteration Plot)\n\n")
  # Plot 1: Shows how balance metrics change across iterations
  print(plot(twang_model, plots = 1))
  cat("\n\n")

  cat("#### Weight Distribution (Boxplot)\n\n")
  # Plot 2: Shows distribution of weights
  print(plot(twang_model, plots = 2))
  cat("\n\n")

  cat("#### Covariate Balance (Standardized Mean Differences)\n\n")
  # Plot 3: Shows balance for each covariate
  print(plot(twang_model, plots = 3))
  cat("\n\n")

  cat("#### Love Plot\n\n")
  # Use cobalt for better visualization
  love_plot <- love.plot(
    bal.tab(ps_formula, 
            data = data_with_weights,
            weights = "weights",
            estimand = "ATT",
            s.d.denom = "treated"),
    thresholds = c(m = 0.1),
    title = paste("Covariate Balance - Imputation", i)
  )
  print(love_plot)
  cat("\n\n")

  cat("#### Weight Histogram\n\n")
  # Plot 6: Histogram of weight proportions
  print(plot(twang_model, plots = 6))
  cat("\n\n")

  cat("#### Balance Summary Table\n\n")
  summary(twang_model) |> print() |> kable()
  cat("\n\n")
}
```

### 2025-08-19 reanalysis-4-outcome_analysis.qmd
analysis audit

# Additional packages
library(emmeans)
library(margins)
library(survey)
library(ggplot2)
library(gridExtra)
library(knitr)
library(tidyr)
library(dplyr)
library(mitools)
library(mice)

# ANALYSIS MODE: Use same as imputation/PS
analysis_mode <- "test"  # Should match imputation and PS modes
```

**Analysis mode:** `r analysis_mode`

@@

For each parameter θ, we combine estimates across m imputations:

@@

5. **Degrees of freedom** (Barnard-Rubin adjustment for finite samples):
   - $df = \frac{1}{\frac{1}{df_{old}} + \frac{1}{df_{obs}}}$
   - Where $df_{old} = (m-1)/r^2$ and $df_{obs} = df_{complete}(1-\lambda)$
   - $r$ = relative increase in variance due to missing data
   - $\lambda$ = fraction of missing information

@@

- Standard errors from `svyglm` already account for IPTW weights
- Pooling preserves the design-based variance estimation
- Joint tests use Li-Raghunathan-Rubin method for χ² statistics

@@

```{r load-data}
# Load the imputed datasets
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))
n_imputations <- length(imputed_datasets)

# Load the twang PS results
ps_results <- readRDS(file.path(reanalysis_data_dir, "ps_results_twang.rds"))

# Extract weights for each imputation
for (i in 1:n_imputations) {
  imputed_datasets[[i]]$iptw_weight <- ps_results$twang_results[[i]]$weights
}

# Summary of first dataset for verification
first_data <- imputed_datasets[[1]]

@@

**DR Severity:**
```{r dr-severity-dist}
table(imputed_datasets[[1]]$dr_severity)
```

@@

**Lapse Status:**
```{r lapse-dist}
table(imputed_datasets[[1]]$ever_lapse_binary)
```

**Three-way Cross-tabulation:**
```{r crosstab}
with(imputed_datasets[[1]], 
     ftable(dr_severity, any_treatment, ever_lapse_binary))
```

@@

```{r fit-models}
# Store models for each imputation
models <- list()
model_summaries <- list()

# Track model fitting progress
model_progress <- data.frame(
  Imputation = integer(),
  Converged = logical(),
  ThreeWay_Terms = integer()
)

@@

  # Create weighted survey design
  design_weighted <- svydesign(
    ids = ~e_mrn_deidentified,
    data = imputed_datasets[[i]],
    weights = ~iptw_weight
  )

  # Fit the three-way interaction model
  model <- svyglm(
    formula = interaction_formula,
    design = design_weighted,
    family = quasibinomial()
  )

@@

  # Extract key coefficients for summary
  coef_summary <- summary(model)$coefficients

@@

## Pool Results Using Rubin's Rules for Survey-Weighted Analysis

```{r pool-results}
# Function to properly pool survey-weighted results using Rubin's rules
pool_survey_results <- function(models, n_imputations) {

  # Extract coefficients and variance-covariance matrices
  all_coefs <- lapply(models, coef)
  all_vcovs <- lapply(models, vcov)

  # Get all coefficient names from first model
  coef_names <- names(all_coefs[[1]])

  # Initialize storage
  pooled_results <- data.frame(
    Term = coef_names,
    Estimate = numeric(length(coef_names)),
    SE = numeric(length(coef_names)),
    df = numeric(length(coef_names)),
    stringsAsFactors = FALSE
  )

  # Apply Rubin's rules for each coefficient
  for (j in 1:length(coef_names)) {
    coef_name <- coef_names[j]

    # Extract coefficient values across imputations
    coef_values <- sapply(all_coefs, function(x) x[coef_name])

    # Step 1: Calculate pooled estimate (Q-bar)
    Q_bar <- mean(coef_values)

    # Step 2: Within-imputation variance (U-bar)
    U_bar <- mean(sapply(all_vcovs, function(x) x[coef_name, coef_name]))

    # Step 3: Between-imputation variance (B)
    B <- var(coef_values)

    # Step 4: Total variance (T) using Rubin's formula
    # T = U_bar + (1 + 1/m) * B
    T <- U_bar + (1 + 1/n_imputations) * B

    # Step 5: Degrees of freedom using Barnard-Rubin adjustment
    # For survey data, we need the complete-data degrees of freedom
    n_obs <- nrow(models[[1]]$data)
    n_params <- length(coef_names)
    df_complete <- n_obs - n_params  # Complete data df

    # Relative increase in variance due to missing data
    r <- (1 + 1/n_imputations) * B / U_bar

    # Fraction of missing information
    lambda <- (r + 2/(df_complete + 3))/(r + 1)

    # Barnard-Rubin adjusted degrees of freedom
    df_old <- (n_imputations - 1) / (r^2)  # Old formula
    df_observed <- df_complete * (1 - lambda)  # Observed data df
    df_adjusted <- 1 / (1/df_old + 1/df_observed)  # Barnard-Rubin adjustment

    # Store results
    pooled_results$Estimate[j] <- Q_bar
    pooled_results$SE[j] <- sqrt(T)
    pooled_results$df[j] <- df_adjusted
  }

  return(pooled_results)
}

# Apply the pooling function
pooled_results <- pool_survey_results(models, n_imputations)

# Calculate additional statistics with proper df
pooled_results <- pooled_results |>
  mutate(
    t_stat = Estimate / SE,
    # Use t-distribution with adjusted df for more accurate p-values
    P_Value = 2 * pt(-abs(t_stat), df = df),
    # Use t-distribution for confidence intervals
    CI_Lower = Estimate - qt(0.975, df = df) * SE,
    CI_Upper = Estimate + qt(0.975, df = df) * SE,
    OR = exp(Estimate),
    OR_CI_Lower = exp(CI_Lower),
    OR_CI_Upper = exp(CI_Upper)
  )

# Display diagnostic information
message("Pooling diagnostics:")
message("  - Number of imputations: ", n_imputations)
message("  - Number of parameters: ", nrow(pooled_results))
message("  - Average degrees of freedom: ", round(mean(pooled_results$df), 1))

# Calculate fraction of missing information for key terms
key_terms <- grep("ever_lapse", pooled_results$Term, value = TRUE)[1:3]
for (term in key_terms) {
  if (term %in% pooled_results$Term) {
    row <- pooled_results[pooled_results$Term == term, ]
    message("  - ", substr(term, 1, 30), "... df = ", round(row$df, 1))
  }
}

@@

```{r joint-tests}
# Function to pool chi-square test statistics across imputations
pool_chi_square_tests <- function(models, hypothesis_strings, test_name) {

@@

# Use the first model to get term names (should be same across all imputations)
first_model <- models[[1]]

@@

  # Increase reference grid limit for large models
  emm_options(rg.limit = 500000)

  # Store contrasts from each imputation
  all_contrasts <- list()

@@

    # Calculate marginal means for all combinations
    emm <- emmeans(models[[i]], 
                   ~ ever_lapse_binary | dr_severity * any_treatment,
                   weights = "proportional")

    # Get contrasts: effect of lapse within each DR × treatment combination
    lapse_contrasts <- contrast(emm, method = "revpairwise", 
                               by = c("dr_severity", "any_treatment"))

@@

  # Pool the contrasts
  # Get unique combinations
  first_df <- all_contrasts[[1]]
  n_contrasts <- nrow(first_df)

>>>>

# Additional packages
library(emmeans)
library(margins)
library(survey)
library(ggplot2)
library(gridExtra)
library(knitr)
library(tidyr)
library(dplyr)
library(mitools)  # For pooling multiple imputation results
library(mice)

# ANALYSIS MODE: Use same as imputation/PS
analysis_mode <- "test"  # Should match imputation and PS modes

# Set parameters based on mode
if (analysis_mode == "test") {
  max_imputations_to_process <- 2  # Process only 2 imputations in test mode
} else {
  max_imputations_to_process <- NULL  # Process all available imputations
}
```

**Analysis mode:** `r analysis_mode`  
**Imputations to process:** `r if(!is.null(max_imputations_to_process)) max_imputations_to_process else "all available"`

@@

Following Rubin (1987) and Barnard & Rubin (1999), for each parameter θ, we combine estimates across m imputations:

@@

5. **Degrees of freedom** (Barnard-Rubin adjustment for finite samples):
   - $r = (1 + \frac{1}{m}) \frac{B}{\bar{U}}$ (relative increase in variance)
   - $df_{old} = (m-1) \times (1 + \frac{1}{r})^2$ (large sample df)
   - $\lambda = \frac{r + 2/(df_{com} + 3)}{r + 1}$ (fraction of missing information)
   - $df_{obs} = \frac{df_{com} + 1}{df_{com} + 3} \times df_{com} \times (1 - \lambda)$ (observed df)
   - $df_{adj} = \frac{df_{old} \times df_{obs}}{df_{old} + df_{obs}}$ (adjusted df)

Where $df_{com}$ is the complete-data degrees of freedom.

**References:**
- Rubin, D.B. (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.
- Barnard, J. & Rubin, D.B. (1999). Small-sample degrees of freedom with multiple imputation. *Biometrika*, 86(4), 948-955.

### Implementation Using mitools

The `mitools` package (Lumley, 2020) provides robust implementations of Rubin's rules and extensions for survey data:

- **`MIcombine()`**: Implements standard Rubin's rules with Barnard-Rubin adjustment
- **`with.imputationList()`**: Applies analyses across imputations
- **`pool.scalar()`**: Combines scalar quantities (e.g., R²)
- **`MIextract()`**: Extracts results from fitted models

This analysis uses `mitools` rather than manual implementation to ensure accuracy and handle edge cases properly.

@@

- Standard errors from `svyglm` already account for IPTW weights
- Pooling preserves the design-based variance estimation
- Joint tests use Li-Raghunathan-Rubin method for χ² statistics (Li et al., 1991)

@@

```{r load-data}
# Load the imputed datasets
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))
n_available <- length(imputed_datasets)

# Load the twang PS results
ps_results <- readRDS(file.path(reanalysis_data_dir, "ps_results_twang.rds"))
n_ps_available <- length(ps_results$twang_results)

# Determine how many imputations to actually process
if (!is.null(max_imputations_to_process)) {
  n_imputations <- min(max_imputations_to_process, n_available, n_ps_available)
  # Subset the datasets if needed
  if (n_imputations < n_available) {
    imputed_datasets <- imputed_datasets[1:n_imputations]
  }
} else {
  n_imputations <- min(n_available, n_ps_available)
}

# Extract weights for each imputation
for (i in 1:n_imputations) {
  imputed_datasets[[i]]$iptw_weight <- ps_results$twang_results[[i]]$weights
}
```

**Available imputed datasets:** `r n_available`  
**Available PS results:** `r n_ps_available`  
**Processing:** `r n_imputations` **datasets** `r if(analysis_mode == "test") "(test mode limit)" else "(all available)"`

```{r continue-load}
# Summary of first dataset for verification
first_data <- imputed_datasets[[1]]

@@

**DR Severity:**
```{r dr-severity-dist}
table(imputed_datasets[[1]]$dr_severity) |> kable()
```

@@

**Lapse Status:**
```{r lapse-dist}
table(imputed_datasets[[1]]$ever_lapse_binary) |> kable()
```

**Three-way Cross-tabulation:**
```{r crosstab}
with(imputed_datasets[[1]], 
     ftable(dr_severity, any_treatment, ever_lapse_binary)) |> kable()
```

@@

```{r fit-models}
models <- list()
model_summaries <- list()

model_progress <- data.frame(
  Imputation = integer(),
  Converged = logical(),
  ThreeWay_Terms = integer()
)

@@

  design_weighted <- svydesign(
    ids = ~e_mrn_deidentified,
    data = imputed_datasets[[i]],
    weights = ~iptw_weight
  )

  model <- svyglm(
    formula = interaction_formula,
    design = design_weighted,
    family = quasibinomial()
  )

@@

  coef_summary <- summary(model)$coefficients

@@

## Pool Results Using Rubin's Rules

```{r pool-results}
# Use mitools for robust pooling implementation
# MIcombine properly implements Rubin's rules with Barnard-Rubin adjustment
pooled_model <- MIcombine(three_way_models)

# Extract pooled results
pooled_summary <- summary(pooled_model)

# Get pooled variance-covariance matrix for later use
pooled_vcov <- vcov(pooled_model)

pooled_results <- data.frame(
  Term = rownames(pooled_summary),
  Estimate = pooled_summary[, "results"],
  SE = pooled_summary[, "se"],
  statistic = pooled_summary[, "(lower"],  
  df = pooled_summary[, "df"],
  p.value = pooled_summary[, "upper)"],
  CI_Lower = pooled_summary[, "2.5 %"],
  CI_Upper = pooled_summary[, "97.5 %"],
  OR = exp(pooled_summary[, "results"]),
  OR_CI_Lower = exp(pooled_summary[, "2.5 %"]),
  OR_CI_Upper = exp(pooled_summary[, "97.5 %"])
)

# Clean column names if needed
names(pooled_results) <- c("Term", "Estimate", "SE", "t_stat", "df", "P_Value", 
                           "CI_Lower", "CI_Upper", "OR", "OR_CI_Lower", "OR_CI_Upper")
# Ensure proper formatting
rownames(pooled_results) <- NULL

### Pooling Diagnostics

```{r pooling-diagnostics}
pool_diag <- data.frame(
  Metric = c("Number of imputations", 
             "Number of parameters",
             "Average degrees of freedom"),
  Value = c(n_imputations,
            nrow(pooled_results),
            round(mean(pooled_results$df), 1))
)

kable(pool_diag, caption = "Pooling diagnostics")

key_terms <- grep("ever_lapse", pooled_results$Term, value = TRUE)[1:min(3, length(grep("ever_lapse", pooled_results$Term, value = TRUE)))]
key_df <- pooled_results[pooled_results$Term %in% key_terms, c("Term", "df")]
key_df$Term <- substr(key_df$Term, 1, 40)
key_df$df <- round(key_df$df, 1)

kable(key_df, caption = "Degrees of freedom for key terms")

@@

```{r joint-tests}
pool_chi_square_tests <- function(models, hypothesis_strings, test_name) {

@@

first_model <- models[[1]]

@@

  emm_options(rg.limit = 500000)

  all_contrasts <- list()

@@

    emm <- emmeans(models[[i]], 
                   ~ ever_lapse_binary | dr_severity * any_treatment,
                   weights = "proportional")

    lapse_contrasts <- contrast(emm, method = "revpairwise", 
                               by = c("dr_severity", "any_treatment"))

@@

  first_df <- all_contrasts[[1]]
  n_contrasts <- nrow(first_df)
